Numerical simulations of the Cordilleran ice sheet through the last glacial cycle

### 2016-01-03 reviews/response-marshall.tex

% ----------------------------------------------------------------------
% Interactive comment text begins
% ----------------------------------------------------------------------

\newcommand{\sechead}[1]{\bigskip\noindent\textbf{#1}}
\newcommand{\referee}[1]{\bigskip\textcolor{journalname}{\textit{#1}}}
\newcommand{\msquote}[1]{\begin{quote}\textit{#1}\end{quote}}

To S.~J.~Marshall,

We thank you very much for this positive review of our manuscript.

@@

% ----------------------------------------------------------------------

\newpage
\sechead{1 \quad Specific comments}

\referee{%
    The Cordilleran Ice Sheet is difficult to model due to its complex
    topography and multiple inception centres (and possibly multiple
    domes/divides), strong regional climatic gradients, which require
    relatively high-resolution climate input fields, and a dearth of
    paleoclimate proxies for western North America to inform spatial and
    temporal variations in climate conditions during the glacial period. The
    authors confront these challenges well, with a adequate ice sheet model
    resolution (5 to 10\,km) and ice physics, carefully calibrated `control'
    climatology \citep[published in][]{Seguinot.etal.2014}), and a good
    exploration of different paleoclimate time series histories in this
    contribution.}

Thank you for summarizing so well the most challenging aspects of our study.

@@

\referee{%
    Nevertheless, it is not clear that ice-core based paleoclimate proxies from
    Greenland or Antarctica are appropriate for western North America. This may
    be particularly true of Greenland proxies, where the amplitude of D-O
    (millennial) climate variations is exceptionally strong and is likely to be
    regional. Because these remote ice core records are `scaled' based on only
    one constraint, producing a CIS maximum configuration that resembles the
    geological record, it is difficult to assess the pre-LGM simulations or the
    details of the modelled ice divide structure, ice thickness, etc. The
    robustness of the conclusion that Greenland and Antarctic ice core records
    are good proxies of glacial climate variability in western North America is
    therefore not so clear, but it is admittedly hard to do better at this
    time. I do wonder if there is any hope from more regional climate proxies
    such as the Logan ice cores or the off-shore Vancouver Island sediment
    records that are cited from Cosma et al. This is worth a short discussion.}

Thank you for raising this point. Choosing proxy records to drive the model
was not trivial and, realising that have not explained this choice in the
first version of the manuscript, we are happy to do it here.

First, we focus on records that have been recognized as proxies for
palaeo-temperature. This includes oxygen isotope (\chem{\delta^{18}O}) records
from the Greenland and Antarctic ice sheet. The Mount Logan ice core
\chem{\delta^{18}O} record has been interpreted as a proxy for source region
rather than palaeotemperature and therefore does not enter this category
\citep{Fisher.etal.2004, Fisher.etal.2008}. Besides ice core records, we use
alkenone unsaturation index (\chem{U^{K'}_{37}}) series from oceanic sediment
cores, a known proxy for sea-surface temperatures \citep{Prahl.Wakeham.1987,
Prahl.etal.1988,Muller.etal.1998}. We do not use $\delta^{18}O$ records from
benthic foraminifera, which have been evaluated worldwide
\citep{Lisiecki.Raymo.2005}, including for core~MD02-2496 near Vancouver Island
\citep{Cosma.etal.2008}, as they are commonly interpreted as proxies for ice
volume on land \citep{Shackleton.1967} rather than temperatures.

Second, we choose to use only records that span over the last 120\,ka in order
to avoid spinning-up the model using a different record than the one tested.
This is a second reason for not including the Mount Logan \chem{\delta^{18}O}
record \citep{Fisher.etal.2004}. Offshore Vancouver Island,
sea-surface temperatures have been reconstructed for core~JT96-09 from alkenone
unsaturation indices over the last 16 ka \citep{Kienast.McKay.2001}, and for nearby
core~MD02-2496 from Mg/Ca ratio in planktonic foraminifera (\emph{N. pachyderma}
and \emph{G. bulloides}) between 12 and 21\,\unit{\chem{^{14}C}\,cal\,ka}
\citep{Taylor.etal.2014}. More recently, the planktonic foraminifera record has
been extended to 10 to ca.~50\,\unit{\chem{^{14}C}\,cal\,ka}
\citep{Taylor.etal.2015}, which could perhaps be used to model Cordilleran ice
sheet growth and decay during the Marine Oxygen Isotope Stage (MIS) 2, but
coult not be directly compared with other runs without spinning-up the model
using a different proxy record.

A shortened version of the preceding explanation has been added to the climate
forcing section of the manuscript. We have also clarified in our conclusions
that we do not consider the GRIP and EPICA ice core records, our preferred
climate drivers, as adequate proxies of climate variability in western North
America, and highlight the need of more regional proxy records to understand
the dynamics of the Cordilleran ice sheet through the last glacial cycle.

### 2016-01-03 reviews/response-marshall.tex

A shortened version of the preceding explanation has been added to the climate
forcing section of the manuscript. We have also clarified in our conclusions
that we do not consider the GRIP and EPICA ice core records, our preferred
climate drivers, as adequate proxies of climate variability in western North
America, and highlight the need of more regional proxy records to understand
the dynamics of the Cordilleran ice sheet through the last glacial cycle.

\referee{%
    Similarly it is difficult to know the errors and uncertainties associated
    with the assumption of fixed modern-day spatial patterns for temperature
    and precipitation. I suspect that the sensitivity of this assumption far
    exceeds that associated with the different paleoclimate proxies. Such that,
    for example, one could readily imagine different assumptions, such as a
    maritime effect that gives reduced glacial cooling near the coast vs. in
    the interior, that is a stronger effect than the difference between
    different paleo-climate proxies with respect to the timing of LGM, ice
    divide structure, etc. But this is a very reasonable start, what the
    authors have done -- there is always going to be more parameter space to
    explore in future studies. As above, I would perhaps just suggest a small
    discussion of the authors' opinion on this question, the uncertainty or
    possible influence of this assumption of modern-day climate patterns.}

Indeed, we chose to keep modern-day spatial patterns for temperature and
precipitation constant in order to limit the number of degrees of freedom in
our study.

Regarding temperature, it is reasonable to think that the changes were greater
inland than near the coast. However, our simulation already produce an
excessive of ice inland. Including such a temperature continentality gradient
in the model while keeping the precipitation pattern constant would thus cause
an even greater mismatch between model results and the geologically
reconstructed Last Glacial Maximum (LGM) ice margins. We have introduced a
short comment of this effect in Sect.~5.1.2 (previously~4.1.2, Ice
configuration during MIS~2).

However, the mismatch we observe between the modelled and reconstructed LGM ice
margins let us think that the assumption of fixed modern-day precipitation
patterns is more critical than the assumption of fixed modern-day temperature
patterns. During phases of ice sheet growth, the presence of mountain ice caps
on the Coast Mountains likely resulted in a decrease of precipitation inland
(Sect.~5.1.2). Furthermore, the adiabatic warming associated with moisture
depletion in the interior (Sect.~5.1.2) may have counterbalanced the potential
continentality gradient discussed above.

@@

\referee{%
    Several minor points and grammatical corrections are included in the
    attached text. Nothing that will require much thought -- this is a really
    impressive piece of research, overall, and I am hard-pressed to find any
    criticism of it. It is one of the easiest reviews I have ever done.
    Congratulations to the authors and thanks for this fine work.}

We have corrected grammatical mistakes given in the attachment. Thank you for
spotting them. And thank you again for supporting our manuscript with such
enthusiasm!

### 2016-01-03 reviews/response-marshall.tex

\referee{%
    The Cordilleran Ice Sheet is difficult to model due to its complex
    topography and multiple inception centres (and possibly multiple
    domes/divides), strong regional climatic gradients, which require
    relatively high-resolution climate input fields, and a dearth of
    paleoclimate proxies for western North America to inform spatial and
    temporal variations in climate conditions during the glacial period. The
    authors confront these challenges well, with a adequate ice sheet model
    resolution (5 to 10\,km) and ice physics, carefully calibrated `control'
    climatology \citep[published in][]{Seguinot.etal.2014}), and a good
    exploration of different paleoclimate time series histories in this
    contribution.}

Thank you for summarizing so well the most challenging aspects of our study.

\referee{%
    Nevertheless, it is not clear that ice-core based paleoclimate proxies from
    Greenland or Antarctica are appropriate for western North America. This may
    be particularly true of Greenland proxies, where the amplitude of D-O
    (millennial) climate variations is exceptionally strong and is likely to be
    regional. Because these remote ice core records are `scaled' based on only
    one constraint, producing a CIS maximum configuration that resembles the
    geological record, it is difficult to assess the pre-LGM simulations or the
    details of the modelled ice divide structure, ice thickness, etc. The
    robustness of the conclusion that Greenland and Antarctic ice core records
    are good proxies of glacial climate variability in western North America is
    therefore not so clear, but it is admittedly hard to do better at this
    time. I do wonder if there is any hope from more regional climate proxies
    such as the Logan ice cores or the off-shore Vancouver Island sediment
    records that are cited from Cosma et al. This is worth a short discussion.}

Thank you for raising this point. Before explaining our choice of
palaeo-climate proxy records, we would like to clarify that we do not consider
the GRIP and EPICA ice core records, our preferred climate drivers, as
satisfactory proxies of climate variability in western North America. Instead,
we think that more regional proxy records are needed in order to better
understand the dynamics of the Cordilleran ice sheet through the last glacial
cycle, and have updated the conclusions to highlight this fact in the
manuscript.

To choose palaeo-climate proxy records to force the model, we first focused on
records that are recognized as proxies for temperature. This includes oxygen
isotope (\chem{\delta^{18}O}) records from the Greenland and Antarctic ice
sheets. Besides ice core records, we use alkenone unsaturation index
(\chem{U^{K'}_{37}}) series from oceanic sediment cores, a known proxy for
sea-surface temperatures \citep{Prahl.Wakeham.1987, Prahl.etal.1988,
Muller.etal.1998}. However, we do not use $\delta^{18}O$ records from benthic
foraminifera, which have been evaluated worldwide \citep{Lisiecki.Raymo.2005},
including for core~MD02-2496 near Vancouver Island \citep{Cosma.etal.2008}, but
are commonly interpreted as proxies for global ice volume on land rather than
for temperatures \citep{Shackleton.1967}. Second, we chose to use only records
that span over the last 120\,ka in order to avoid spinning-up the model using a
different record than the one being tested. The Mount Logan ice core
\chem{\delta^{18}O} record covers only the last 30\,Ka and has been interpreted
as a proxy for source region rather than for palaeo-temperature
\citep{Fisher.etal.2004, Fisher.etal.2008} and thus do not fit our criteria.

Concerning sediment cores offshore Vancouver Island, sea-surface temperatures
have been reconstructed for core~JT96-09 from alkenone unsaturation indices
over the last 16 ka \citep{Kienast.McKay.2001}, and for the nearby
core~MD02-2496 from the Mg/Ca ratio in planktonic foraminifera (\emph{N.
pachyderma} and \emph{G. bulloides}) between 12 and
21\,\unit{\chem{^{14}C}\,cal\,ka} \citep{Taylor.etal.2014}. Very recently, this
planktonic foraminifera record has been extended to cover the period from 10 to
ca.~50\,\unit{\chem{^{14}C}\,cal\,ka} \citep{Taylor.etal.2015}, and thus it
could perhaps be used to model Cordilleran ice sheet growth and decay during
the Marine Oxygen Isotope Stage (MIS) 2, but such a simulation could not be
directly compared with other runs without spinning-up the model using a
different proxy record.

A shortened version of the preceding explanation has been added to the climate
forcing section of the manuscript.

### 2016-01-03 reviews/response-marshall.tex

A shortened version of the preceding explanation has been added to the climate
forcing section of the manuscript.

\referee{%
    Similarly it is difficult to know the errors and uncertainties associated
    with the assumption of fixed modern-day spatial patterns for temperature
    and precipitation. I suspect that the sensitivity of this assumption far
    exceeds that associated with the different paleoclimate proxies. Such that,
    for example, one could readily imagine different assumptions, such as a
    maritime effect that gives reduced glacial cooling near the coast vs. in
    the interior, that is a stronger effect than the difference between
    different paleo-climate proxies with respect to the timing of LGM, ice
    divide structure, etc. But this is a very reasonable start, what the
    authors have done -- there is always going to be more parameter space to
    explore in future studies. As above, I would perhaps just suggest a small
    discussion of the authors' opinion on this question, the uncertainty or
    possible influence of this assumption of modern-day climate patterns.}

Indeed, we chose to keep modern-day spatial patterns for temperature and
precipitation constant in order to limit the number of degrees of freedom in
our study.

Regarding temperature, it is reasonable to think that the changes were greater
inland than near the coast. However, our simulations already produce an
excess of ice inland. Including such a temperature continentality gradient
in the model while keeping the precipitation pattern constant would thus cause
an even greater mismatch between the model results and the geologically
reconstructed Last Glacial Maximum (LGM) ice margins. We have introduced a
short comment of this effect in Sect.~5.1.2 (previously~4.1.2, Ice
configuration during MIS~2).

@@

However, the mismatch we observe between the modelled and reconstructed LGM ice
margins let us think that the assumption of fixed modern-day precipitation
patterns is more critical than the assumption of fixed modern-day temperature
patterns. During phases of ice sheet growth, the presence of mountain ice caps
on the Coast Mountains likely resulted in a decrease of precipitation inland
(Sect.~5.1.2). Furthermore, the adiabatic warming associated with moisture
depletion in the interior (Sect.~5.1.2) may have counterbalanced the potential
continentality gradient discussed above.

\referee{%
    Several minor points and grammatical corrections are included in the
    attached text. Nothing that will require much thought -- this is a really
    impressive piece of research, overall, and I am hard-pressed to find any
    criticism of it. It is one of the easiest reviews I have ever done.
    Congratulations to the authors and thanks for this fine work.}

We have corrected grammatical mistakes given in the attachment. Thank you for
spotting them. And thank you again very much for supporting our manuscript with
such enthusiasm!

### 2016-01-04 reviews/response-marshall.tex

Thank you for summarizing so well the most challenging aspects of our study.

\referee{%
    Nevertheless, it is not clear that ice-core based paleoclimate proxies from
    Greenland or Antarctica are appropriate for western North America. This may
    be particularly true of Greenland proxies, where the amplitude of D-O
    (millennial) climate variations is exceptionally strong and is likely to be
    regional. Because these remote ice core records are `scaled' based on only
    one constraint, producing a CIS maximum configuration that resembles the
    geological record, it is difficult to assess the pre-LGM simulations or the
    details of the modelled ice divide structure, ice thickness, etc. The
    robustness of the conclusion that Greenland and Antarctic ice core records
    are good proxies of glacial climate variability in western North America is
    therefore not so clear, but it is admittedly hard to do better at this
    time. I do wonder if there is any hope from more regional climate proxies
    such as the Logan ice cores or the off-shore Vancouver Island sediment
    records that are cited from Cosma et al. This is worth a short discussion.}

Thank you for raising this point. Before explaining our choice of
palaeo-climate proxy records, we would like to clarify that we do not consider
the GRIP and EPICA ice core records, our preferred climate drivers, as
satisfactory proxies of climate variability in western North America. Instead,
we think that more regional proxy records are needed in order to better
understand the dynamics of the Cordilleran ice sheet through the last glacial
cycle, and have updated the conclusions to highlight this fact in the
manuscript.

To choose palaeo-climate proxy records to force the model, we first focused on
records that are recognized as proxies for temperature. These include oxygen
isotope (\chem{\delta^{18}O}) records from the Greenland and Antarctic ice
sheets. Besides ice core records, we use alkenone unsaturation index
(\chem{U^{K'}_{37}}) series from oceanic sediment cores, a known proxy for
sea-surface temperatures \citep{Prahl.Wakeham.1987, Prahl.etal.1988,
Muller.etal.1998}. However, we do not use $\delta^{18}O$ records from benthic
foraminifera, which have been evaluated worldwide \citep{Lisiecki.Raymo.2005},
including for core~MD02-2496 near Vancouver Island \citep{Cosma.etal.2008}, but
are commonly interpreted as proxies for global ice volume on land rather than
for temperatures \citep{Shackleton.1967}. Second, we chose to use only records
that span over the last 120\,ka in order to avoid spinning-up the model using a
different record than the one being tested. The Mount Logan ice core
\chem{\delta^{18}O} record covers only the last 30\,Ka and has been interpreted
as a proxy for source region rather than for palaeo-temperature
\citep{Fisher.etal.2004, Fisher.etal.2008} and thus does not fit our criteria.

Concerning sediment cores offshore Vancouver Island, sea-surface temperatures
have been reconstructed for core~JT96-09 from alkenone unsaturation indices
over the last 16 ka \citep{Kienast.McKay.2001}, and for the nearby
core~MD02-2496 from the Mg/Ca ratio in planktonic foraminifera (\emph{N.
pachyderma} and \emph{G. bulloides}) between 12 and
21\,\unit{\chem{^{14}C}\,cal\,ka} \citep{Taylor.etal.2014}. Very recently, this
planktonic foraminifera record has been extended to cover the period from 10 to
ca.~50\,\unit{\chem{^{14}C}\,cal\,ka} \citep{Taylor.etal.2015}, and thus it
could perhaps be used to model the growth and decay of the Cordilleran ice
sheet during
the Marine Oxygen Isotope Stage (MIS) 2, but such a simulation could not be
directly compared with other runs without spinning-up the model using a
different proxy record.

### 2016-01-04 reviews/response-jarosch.tex

By ``derived'' we meant that the ETOPO1 data is simply re-sampled with linear
interpolation from the original to the projected grid. We have clarified this in
the methods.

Thank you for pointing out the study by \citet{Clarke.etal.2013}.
Indeed, the ETOPO1 data contains present day ice volumes. The most problematic
part of the model domain in this respect is by far that of the Wrangell and
St.-Elias mountains where ice thicknesses up to 1200\,m have recently been
measured by low-frequency radar \citep{Rignot.etal.2013}. In this area, located
over the USA Canada border and just north of 60{\degree}N and thus not included
in the study by \citet{Clarke.etal.2013}, temperate ice, surge dynamics
and deep subglacial depressions in the icefield interior pose a fundamental
challenge to reconstructing basal topography for the entire ice cap. Although
it is clear that our model overestimates ice thickness in this region, we are
not aware of bed topography data or reconstructions that could be used to force
the ice sheet model. This said, with the exception of the Wrangell and
St.-Elias mountains ice field, present-day ice volumes are small relative to
the ice volumes concerned in our study.

We have added a description of the limitations of ETOPO1 data in the methods,
and a comment on overestimated ice thicknesses in the Wrangell and St.-Elias
mountains in Sect.~5.2.2 (Major ice-dispersal centres).

@@

We agree and have reformulated this statement. Thank you for spotting this.

\referee{%
    \textbf{P4153 L5-6:}
    As stated here, ice rheology within the used ice sheet model is based on
    \citet{Aschwanden.etal.2012}. This enthalpy based formulation has proofed
    itself to be very suitable for estimating ice rheology in ice sheet models,
    but it also depends on several parameters to translate enthalpy within the
    ice to ice viscosity \citep[see][equations 62-65)]{Aschwanden.etal.2012}.
    The authors do not mention any of these parameters (e.g. any of the rate
    factors or nonlinear power $n$ ) within the manuscript or in Table~1. I
    have mentioned above in the general comments section that parameters used
    in ice rheology and basal sliding formulations are important model
    parameters which will influence the ice sheet model output and that a basic
    sensitivity study on those parameters is required to understand the model
    results. Here the authors could start with listing the parameters used in
    the ice rheology formulation, than continue with estimating uncertainties
    for those from literature and afterwards perform additional model
    simulations to identify the influence of the chosen parameter sets on the
    ice volume and ice margin position history the model creates. In the end
    the authors will be able to identify the relative importance of
    uncertainties in driving climate as well as model parameters, which will
    strengthen their discussion in section~4.}

Our changes described above partly address this comment. To make our parameter
choices clear, we have included all default parameter values mentioned
in this response in the manuscript's Table~1 (with the alternative values used
in the sensitivity tests being given in a separate table).

@@

As in \citet{Seguinot.etal.2014}, the NARR data fields have been bilinearly
interpolated from the NARR 32\,km Lambert grid to the computational domain
grids. We apologize for omitting this information and have now added it to the
manuscript.

We called the NARR data ``high resolution'' because its spatial resolution is
higher than most other atmospheric reanalyses. But we agree that it is still
too coarse for ice flow modelling and that temperature and precipitation
downscaling techniques could potentially correct for the errors caused by
the resolution gap. We have rephrased the sentence to highlight this.

### 2016-01-04 reviews/response-jarosch.tex

\sechead{2 \quad Specific Comments}

\referee{%
    I refer to text locations in the discussion paper by page number (P) and
    line number range (L) for the specific comments.}

\referee{%
    \textbf{P4151 L11-16:} In this sentence the authors refer back to their
    previous work \citep{Seguinot.etal.2014} and highlight that the NARR
    temperature and precipitation fields are the most suitable present day
    climate datasets to be used. Especially since the NARR precipitation fields
    include steep precipitation gradients which are required as identified by
    et \citet{Seguinot.etal.2014}. NARR is delivered on a 32\,km Lambert grid,
    and thus it is questionable how ``steep'' these gradients can be, given
    the rather smooth representation of the existing topography on a 32\,km
    grid. \citet{Seguinot.etal.2014} have partly discussed that however. NARR
    precipitation and temperature fields have been evaluated in detail based on
    available station data for large parts of the study domain dealt with in
    this manuscript. This evaluation \citep{Jarosch.etal.2012} demonstrated
    that NARR has difficulties simulating orographic processes in the Coast
    Mountains which in turn results in unrealistic atmospheric conditions over
    the Rocky Mountains. \citet{Jarosch.etal.2012} further concluded that
    physics based downscaling is required to adequately drive glacier models in
    that region. The authors should argue in more detail here why they think
    that NARR precipitation fields at 32\,km are adequate to drive their model
    and reflect their arguments with the findings of \citet{Jarosch.etal.2012}.
    A solid argument here is of special importance as the authors assume the
    present day precipitation fields to be valid throughout their model time
    period (120\,ky to present) without further corrections (cf. section~2.4
    equation~6).}

We admit that the NARR climate forcing used in our simulations has its
limitations.

In our previous study \citep{Seguinot.etal.2014}, we have evaluated the
performance of NARR in forcing constant-climate simulations of the Cordilleran
ice sheet against that of an observation-based data set
\citep[WorldClim]{Hijmans.etal.2005}. Indeed, the use of NARR in these
simulations produced slightly different patterns of glaciation relative to
WorldClim, including more extensive ice cover on the Columbia and Rocky
mountains \citep[Figs.~6--7]{Seguinot.etal.2014}. Our simulations have shown
that these differences are mainly caused by disparities in the precipitation
fields of the two data sets \citep[Figs.13--14]{Seguinot.etal.2014}, which may,
indeed, be related to the fact that the NARR has difficulties with simulating
orographic processes in some areas of steep topography as discussed by
\citet{Jarosch.etal.2012}.

In the light of your comments, we have introduced a discussion of the NARR and
its limitations in Sect.~2.6 (Climate forcing) and in Sect.~5.1.2 (Ice
configuration during MIS~2). We acknowledge that the implementation of the
downscaling method presented by \citet{Jarosch.etal.2012} may address these
limitations. However, extending this downscaling method to the entire the model
domain used in our study could be challenging, because the northern part of the
model domain is characterized by stronger precipitation gradients and includes
much fewer weather stations than the computational domain of
\citet{Jarosch.etal.2012}.

### 2016-01-04 reviews/response-jarosch.tex

Thank you for spotting these two typos. We have corrected them.

\referee{%
    I hope the authors find my comments helpful in revising their manuscript
    and wish them success for their future endeavours.}

% ----------------------------------------------------------------------

\begin{thebibliography}{69}

\bibitem[{Aschwanden et~al.(2012)Aschwanden, Bueler, Khroulev, and
  Blatter}]{Aschwanden.etal.2012}
Aschwanden, A., Bueler, E., Khroulev, C., and Blatter, H.: An enthalpy
  formulation for glaciers and ice sheets, J. Glaciol., 58, 441--457,
  \doi{10.3189/2012JoG11J088}, 2012.

\bibitem[{Aschwanden et~al.(2013)Aschwanden, A{\dh}algeirsd\'{o}ttir, and
  Khroulev}]{Aschwanden.etal.2013}
Aschwanden, A., A{\dh}algeirsd\'{o}ttir, G., and Khroulev, C.: Hindcasting to
  measure ice sheet model sensitivity to initial states, The Cryosphere, 7,
  1083--1093, \doi{10.5194/tc-7-1083-2013}, 2013.

\bibitem[{Bueler and Brown(2009)}]{Bueler.Brown.2009}
Bueler, E. and Brown, J.: Shallow shelf approximation as a ``sliding law'' in
  a~thermodynamically coupled ice sheet model, J. Geophys. Res., 114, F03\,008,
  \doi{10.1029/2008JF001179}, 2009.

\bibitem[{Bueler and van Pelt(2015)}]{Bueler.Pelt.2015}
Bueler, E. and van Pelt, W.: Mass-conserving subglacial hydrology in the
  Parallel Ice Sheet Model version 0.6, Geosci. Model Dev., 8, 1613--1635,
  \doi{10.5194/gmd-8-1613-2015}, 2015.

\bibitem[{Clague(1981)}]{Clague.1981}
Clague, J.~J.: Late Quaternary Geology and Geochronology of British Columbia
  Part 2: Summary and Discussion of Radiocarbon-Dated Quaternary History, Geol.
  Surv. of Can., Ottawa, ON, Paper 80-35, \doi{10.4095/119439}, 1981.

\bibitem[{Clarke et~al.(2013)Clarke, Anslow, Jarosch, Radi{\'{c}}, Menounos,
  Bolch, and Berthier}]{Clarke.etal.2013}
Clarke, G. K.~C., Anslow, F.~S., Jarosch, A.~H., Radi{\'{c}}, V., Menounos, B.,
  Bolch, T., and Berthier, E.: Ice Volume and Subglacial Topography for Western
  Canadian Glaciers from Mass Balance Fields, Thinning Rates, and a Bed Stress
  Model, J. Climate, 26, 4282--4303, \doi{10.1175/jcli-d-12-00513.1}, 2013.

\bibitem[{Cuffey and Paterson(2010)}]{Cuffey.Paterson.2010}
Cuffey, K.~M. and Paterson, W. S.~B.: The physics of glaciers, Elsevier,
  Amsterdam, 2010.

\bibitem[{Glen(1952)}]{Glen.1952}
Glen, J.: Experiments on the deformation of ice, J. Glaciol., 2, 111--114,
  1952.

\bibitem[{Greve(1997)}]{Greve.1997}
Greve, R.: A continuum-mechanical formulation for shallow polythermal ice
  sheets, Philos. T. R. Soc. A, 355, 921--974, \doi{10.1098/rsta.1997.0050},
  1997.

\bibitem[{Hijmans et~al.(2005)Hijmans, Cameron, Parra, Jones, and
  Jarvis}]{Hijmans.etal.2005}
Hijmans, R.~J., Cameron, S.~E., Parra, J.~L., Jones, P.~G., and Jarvis, A.:
  Very high resolution interpolated climate surfaces for global land areas,
  Int. J. Climatol., 25, 1965--1978, \doi{10.1002/joc.1276}, 2005.

\bibitem[{Jarosch et~al.(2012)Jarosch, Anslow, and Clarke}]{Jarosch.etal.2012}
Jarosch, A.~H., Anslow, F.~S., and Clarke, G. K.~C.: High-resolution
  precipitation and temperature downscaling for glacier models, Clim. Dynam.,
  38, 391--409, \doi{10.1007/s00382-010-0949-1}, 2012.

\bibitem[{Lisiecki and Raymo(2005)}]{Lisiecki.Raymo.2005}
Lisiecki, L.~E. and Raymo, M.~E.: A Pliocene-Pleistocene stack of 57 globally
  distributed benthic $\delta^{18}$O records, Paleoceanography, 20, PA1003,
  \doi{10.1029/2004pa001071}, 2005.

\bibitem[{Lliboutry and Duval(1985)}]{Lliboutry.Duval.1985}
Lliboutry, L.~A. and Duval, P.: Various isotropic and anisotropic ices found in
  glaciers and polar ice caps and their corresponding rheologies, Ann.
  Geophys., 3, 207--224, 1985.

\bibitem[{L{\"u}thi et~al.(2002)L{\"u}thi, Funk, Iken, Gogineni, and
  Truffer}]{Luthi.etal.2002}
L{\"u}thi, M., Funk, M., Iken, A., Gogineni, S., and Truffer, M.: Mechanisms of
  fast flow in {J}akobshavns {I}sbr{\ae}, {G}reenland; {P}art {III}:
  measurements of ice deformation, temperature and cross-borehole conductivity
  in boreholes to the bedrock, J. Glaciol., 48, 369--385,
  \doi{10.3189/172756502781831322}, 2002.

\bibitem[{Martin et~al.(2011)Martin, Winkelmann, Haseloff, Albrecht, Bueler,
  Khroulev, and Levermann}]{Martin.etal.2011}
Martin, M.~A., Winkelmann, R., Haseloff, M., Albrecht, T., Bueler, E.,
  Khroulev, C., and Levermann, A.: The Potsdam Parallel Ice Sheet Model
  (PISM-PIK) -- Part 2: Dynamic equilibrium simulation of the Antarctic ice
  sheet, The Cryosphere, 5, 727--740, \doi{10.5194/tc-5-727-2011}, 2011.

\bibitem[{Nye(1953)}]{Nye.1953}
Nye, J.~F.: The Flow Law of Ice from Measurements in Glacier Tunnels,
  Laboratory Experiments and the Jungfraufirn Borehole Experiment, Proc. R.
  Soc. London, Ser. A, 219, 477--489, 1953.

\bibitem[{Paterson and Budd(1982)}]{Paterson.Budd.1982}
Paterson, W. S.~B. and Budd, W.~F.: Flow parameters for ice sheet modeling,
  Cold Reg. Sci. Technol., 6, 175--177, 1982.

\bibitem[{Rignot et~al.(2013)Rignot, Mouginot, Larsen, Gim, and
  Kirchner}]{Rignot.etal.2013}
Rignot, E., Mouginot, J., Larsen, C.~F., Gim, Y., and Kirchner, D.:
  Low-frequency radar sounding of temperate ice masses in Southern Alaska,
  Geophys. Res. Lett., 40, 5399--5405, \doi{10.1002/2013gl057452}, 2013.

\bibitem[{Seguinot(2014)}]{Seguinot.2014}
Seguinot, J.: Numerical modelling of the Cordilleran ice sheet, Ph.D. thesis,
  Stockholm University,
  \urlprefix\url{http://urn.kb.se/resolve?urn=urn:nbn:se:su:diva-106815}, 2014.

\bibitem[{Seguinot et~al.(2014)Seguinot, Khroulev, Rogozhina, Stroeven, and
  Zhang}]{Seguinot.etal.2014}
Seguinot, J., Khroulev, C., Rogozhina, I., Stroeven, A.~P., and Zhang, Q.: The
  effect of climate forcing on numerical simulations of the {C}ordilleran ice
  sheet at the {L}ast {G}lacial {M}aximum, The Cryosphere, 8, 1087--1103,
  \doi{10.5194/tc-8-1087-2014}, 2014.

\bibitem[{the PISM~authors(2015)}]{PISM-authors.2015}
the PISM~authors: {PISM}, a~{P}arallel {I}ce {S}heet {M}odel,
  \urlprefix\url{http://www.pism-docs.org}, 2015.

\bibitem[{Tulaczyk et~al.(2000)Tulaczyk, Kamb, and
  Engelhardt}]{Tulaczyk.etal.2000}
Tulaczyk, S., Kamb, W.~B., and Engelhardt, H.~F.: Basal mechanics of Ice Stream
  B, west Antarctica: 1. Till mechanics, J. Geophys. Res., 105, 463,
  \doi{10.1029/1999jb900329}, 2000.

\bibitem[{Winkelmann et~al.(2011)Winkelmann, Martin, Haseloff, Albrecht,
  Bueler, Khroulev, and Levermann}]{Winkelmann.etal.2011}
Winkelmann, R., Martin, M.~A., Haseloff, M., Albrecht, T., Bueler, E.,
  Khroulev, C., and Levermann, A.: The {P}otsdam {P}arallel {I}ce {S}heet
  {M}odel ({PISM-PIK}) -- Part 1: model description, The Cryosphere, 5,
  715--726, \doi{10.5194/tc-5-715-2011}, 2011.

\end{thebibliography}
